---
title: "Pathway Crosstalk Visualizations"
author: "Menna Arafat"
date:  "2024-9-25"
output: html_document
---


```{r}
dir.create("output")
```


## Load Required Libraries

```{r}
suppressPackageStartupMessages({
library(plyr)
library(dplyr)
library(igraph)
library(tidyr)
library(readxl)
library(openxlsx)
library(scales)
library(ggplot2)
library(tibble)
library(enrichplot)
library(ggrepel)
library(ggraph)
library(tidygraph)
library(RColorBrewer)
  })
```

## Load Data
Load the crosstalk data and pathway terms/components from Excel files.

```{r}
data = read.xlsx("https://github.com/Menna-Arafat/Pathway-CrossTalk/blob/master/data/crosstalk_results_permutation.xlsx", sheet = "crosstalk")
terms = read.xlsx("https://github.com/Menna-Arafat/Pathway-CrossTalk/blob/master/data/enrichment_top_30_features_factor1_gprofiler.xlsx", sheet = "terms") %>%
  column_to_rownames(".id") %>%
  t() %>%
  as.data.frame()
```

## Filter Significant Pathways
Apply a p-value cutoff of 0.1 to select significant pathways.

```{r}
data = data %>% filter(p.value <= 0.1 & pathway2 != "Disease")
terms = terms[, colnames(terms) %in% c(data$pathway1, data$pathway2)]
```

## Add Transcription Factors

```{r}
TF = read_excel("data/enrichment_top_30_features_factor1_gprofiler.xlsx", sheet= "TF")[c(2,11)]
TF$term_name = TF$term_name %>% lapply(., function(i) strsplit(i, ":|;")[[1]][2])

TF_list = TF %>% 
  column_to_rownames("term_name") %>%
  t() %>%
  as.data.frame() %>%
  as.list() %>%
  lapply(., function(x) strsplit(x, ",")[[1]])
```

## Build Lists of Significant Pathways

```{r}
paths_lists = as.list(terms)
paths_lists = lapply(paths_lists, function(x) x[x != "" & !is.na(x)])
all_lists = c(paths_lists, TF_list)

```

## Convert List to Long Format

```{r}
list2df <- function(inputList) {
  ldf <- lapply(seq_len(length(inputList)), function(i) {
    data.frame(categoryID = rep(names(inputList[i]), length(inputList[[i]])),
               Gene = inputList[[i]])
  })
  ldply(ldf, rbind)
}
```

## Convert List to Graph Object

```{r}
list2graph <- function(inputList, directed = FALSE) {
  x <- list2df(inputList)
  g <- graph_from_data_frame(x, directed = directed)
  return(g)
}
```

## Build Igraph Object

```{r}
graph = list2graph(all_lists)
```

## Set Colors for Igraph

```{r}
palette = colorRampPalette(c("#F6E8C3", "#B0C4DE", "#E3B31C", "tan", "purple", "maroon3", "#21908CFF", "darkgrey"))
palette(40)
colors = palette(length(all_lists))

all_col = rep("lightgrey", length(V(graph)))

for(i in seq_along(paths_lists)) {
  all_col[V(graph)$name %in% names(paths_lists)[i]] = colors[i]
}
all_col[V(graph)$name %in% names(TF_list)] = "#96D73F"
```

## Create Tidygraph Object

```{r}
graph_tbl <- as_tbl_graph(graph) %>%
  activate(nodes) %>%
  mutate(size = ifelse(name %in% names(paths_lists), 6,
                       ifelse(name %in% names(TF_list), 4, 2)),
         label_size = ifelse(name %in% names(paths_lists), 2.5, 2),
         shape = ifelse(name %in% names(TF_list), "2", "1"),
         col = all_col)
```

## Calculate Centroids for Each Community

```{r}
layout <- layout_with_kk(graph_tbl)

centroids <- sapply(seq_along(paths_lists), function(i) {
  nodes = V(graph)$name %in% names(paths_lists[[i]])
  colMeans(layout[nodes, ])
}) %>% t() %>% as.data.frame()

row.names(centroids) = names(paths_lists)
```

## Plot Using Ggraph

```{r}
p = ggraph(graph_tbl, layout = 'circle') +  
  geom_edge_fan(aes(alpha = 0.5), color = "lightgrey", show.legend = FALSE) + 
  geom_node_point(aes(size = size, color = col, shape = shape), show.legend = FALSE) + 
  scale_color_identity() + 
  scale_size_identity() +
  geom_node_text(aes(label = name, size = label_size), show.legend = FALSE, repel = TRUE) +
  theme_graph(fg_text_colour = 'black') +
  theme(element_text(size = 13))
```

## Save the Plot

```{r}
print(p)
ggsave("output/crossTalk_circular.png", p, width = 11, height = 7, dpi = 600)
```

